Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-26 paper.tex
updated more figures

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values are obtained by averaging the predictions for each observation from from all 500
trees. We initially employed a training/test set split (80\% of the dataset was assigned to the
training set) and these results are reported in Table \ref{tab:rfmodel-results}. We also developed
models (Section \ref{sec:an-altern-valid} using the entire dataset and observed similar RMSE and
$R^2$ values. Given the fact that the random forest algorithm builds a model based on out-of-bag
data, using the entire dataset rather than a single training/test split this is not surprising.

In this work, we employed the CDK to evaluate 109 2D and constitutional descriptors for each
individual molecule. For each dataset, we performed objective feature selection by removing
descriptors with constant or near-constant values followed by removal of descriptors that are highly
correlated with others (using an $R^2$ cutoff of 0.8). The sizes ($N_{desc}$) of the final
descriptor pool for each dataset are summarized in Table \ref{tab:data}.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, the most significant cliffs
exhibit significant variance. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$
as the aggregation function for subsequent analyses. Fig.~\ref{fig:rfmodel-orig} summarizes the
performance of the models built using the reported activity data for the Costanzo, Kalla and Dai
datasets. Each of these model employed the entire dataset (rather than splitting into a training and
testing set). These models are relatively poor. This can be partially ascribed to the small sizes of
the dataset. It should be note that this methodology is not aimed at modeling the original, usually
small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
not necessarily lead to very robust or reliable machine learning models. Instead, the methodology
focuses on the \emph{pairwise} dataset -- which are usually large enough for a statistically
significant model.

@@

Since the model is built on pairs of observations, performance measures characterizing the model are
based on the accuracy of pairwise SALI values. The model can also be tested in a slightly different
approach by checking whether it can correctly identify cliffs between a new molecule and the member
of the training set. To test this, we consider the original set of $n$ individual molecules and hold
out $m$ molecules (as opposed to holding out a set of observations from the pairwise dataset). We
then evaluate the SALI values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest
model. In contrast to the models previously described, we rebuilt these models using the entire set
of $n$ molecules. The model is then used to predict the SALI values between the $m$ holdout
molecules and the $(n-m)$ remaining molecules.

We first performed this analysis with the Kalla dataset. Since predictions of new molecules involve
$n$ predictions per molecule, we removed three molecules from the dataset and rebuilt the random
forest model. The hold out molecules, were specifically selected since they displayed significant
activity cliffs with various members of the training set. We then evaluated the pairwise descriptor
values for these three with the training set molecules and obtained predictions of the log(SALI)
values. Fig.~\ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
as well as for the prediction set. From a numerical point of view, the performance of the model on
he prediction set degrades somewhat (RMSE of 0.51 versus 0.41 for the training set). As noted above,
the variance of the prediction is higher at the lower scale of log(SALI) values. However, at the
higher end of the scale there a number of pairs that are well predicted. Yet, it is clear that a
number of relatively significant activity cliffs have been underestimated.

Given that large SALI values can arise due to high structural similarity, even when the activity
difference is small, simply examining log(SALI) values is not completely informative. Specifically,
we are interested in the predictions for the actual activity cliffs, where both the difference in
activity and the structural similarity is very high. Fig.~\ref{fig:tset-pset-kalla}B visualizes this
information. Each point corresponds to a pair of molecules, which are shaded by the absolute
prediction residual for that pair. Thus, the ``true'' cliffs are represented by points lying towards
the bottom right corner of the plot. In this case we see that there are a number of pairs of
molecules with $1 - T_C \leq 0.1$ and 100-fold or better difference in activity. While many of these
pairs are associated with a high prediction residual, there are a number with low to medium
residuals.

Fig.~\ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
and a selection of structures, which which these hold out molecules were predicted to exhibit an
activity cliff. For each molecule, the $K_i$ value is listed along with the absolute residual value
(log(SALI) units). For molecules \textbf{349288} and \textbf{349699}, the predicted cliffs are
relatively accurate, as evidenced by the low residuals. For \textbf{349138} on the other hand, most
of the residuals were were relatively high. While we have highlighted a number of cliffs,
Fig.~\ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
average, worse than for the other two molecules.

Fig.~\ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
applied to the Dai dataset. As before we removed three molecules (top row of
Fig.~\ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
41 molecules. We then predicted the pairwise SALI values for the three hold-out molecules with the
remaining 41 molecules (i.e., the prediction set). A plot of the predicted versus observed log(SALI)
values for the training and prediction sets are displayed in Fig.~\ref{fig:tset-pset-dai}A. As
before the model performs poorly on the lesser cliffs, but improves for the more significant cliffs
(RMSE for the training set and prediction sets were 0.37 and 0.43 log units respectively). However,
the 2 most significant cliffs are under-estimated. However, in comparison to the Kalla dataset, the
distribution of absolute residuals aggregated by the hold-out molecule
(Fig.~\ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
0.3 log units.  For this dataset, the number of ``real'' cliffs is relatively low as shown in
Fig.~\ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
significant cliffs are relatively well predicted - in fact, the maximum residuals are observed for
the least significant cliffs.

We then considered the three hold-out molecules and some of the members of the training set with
which they are predicted to show activity cliffs. For this dataset, there are relatively few
``true'' activity cliffs. For example, while molecules \textbf{371259} and \textbf{371307} have a
1000-fold difference in activity, their $T_c = 0.68$, and would not represent a truly significant
cliff. However, the pair is well predicted with a residual of 0.04 log units. if we consider the
subset of predictions where the $T_c \geq 0.8$ we note that the median absolute residual is 0.24 log
units. However, the activity differences are not always significant ranging from 1.1-fold to
13-fold. Two such examples are shown in Fig.~\ref{fig:dai-comparison}, where the only difference is
in the position of the methyl substituent. Similar behavior is observed with the other hold-out
molecules.

@@

Given that many of the models show relatively poor statistics on the pairwise data, it is important
to note that much of this is due to the large variance in the predictions for the less significant
cliffs. One possible reason for the increase in variance for these portions of the dataset is the
fact that the distribution of the log(SALI) values tends to be skewed to the right
(Fig.~\ref{fig:logsalidist}). In general we see that the smaller SALI values constitute a relatively
small portion of the dataset.

@@

@@

\ctable[cap={}, caption={The hold out molecules for the Kalla dataset
  and training set members with which the hold outs exhibit predicted
  activity cliffs. Bold numbers are ChEMBL MOLREGNO values and numbers
  in parentheses are the absolute prediction
  residual in log(SALI) units.}, figure, botcap,
label={fig:kalla-comparison}]    
{c}
{}
{\includegraphics[width=\linewidth]{kalla-comparison}}

>>>>

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values are obtained by averaging the predictions for each observation from from all 500
trees. We initially employed a training/test set split (80\% of the dataset was randomly assigned to
the training set) and these results are reported in Table \ref{tab:rfmodel-results}. We also
developed models (Section \ref{sec:an-altern-valid}) using the entire dataset and observed similar
RMSE and $R^2$ values. Given the fact that the random forest algorithm builds a model based on
out-of-bag data, the similarity in performance of models built using the entire dataset and models
built using a training/test split is not surprising.

We employed the CDK to evaluate 109 2D and constitutional descriptors for each  molecule. For each
dataset, we performed objective feature selection by removing descriptors with constant or
near-constant values followed by removal of descriptors that are highly correlated with others
(using an $R^2$ cutoff of 0.8). The sizes ($N_{desc}$) of the final descriptor pool for each dataset
are summarized in Table \ref{tab:data}.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, the most significant cliffs
exhibit significant variance. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$
as the aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and
Figure \ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported
activity data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of
these models employed the entire dataset (rather than splitting into a training and testing set) and
exhibit relatively poor performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units)
and $R^2$ (0.18 to 0.34).. This can be partially ascribed to the small sizes of the dataset. It
should be note that this methodology is not aimed at modeling the original, usually small, SAR
dataset. It is well known (and evident from the current analyses) that small datasets do not
necessarily lead to very robust or reliable machine learning models. Instead, the methodology
focuses on the \emph{pairwise} dataset -- which are usually large enough for a statistically
significant model.

@@

Since the model is built on pairs of observations, performance measures characterizing the model are
based on the accuracy of pairwise SALI values. Thus the preceding section validated models based on
holding out pairs of observations and evaluating the RMSE and $R^2$ for the predicted values of
these held out pairs.

The model can also be tested in a slightly different approach by checking whether it can correctly
identify cliffs between a new molecule and the member of the training set. In fact, such a scenario
would be the anticipated, prospective use case for such a modeling methodology. To test this, we
consider the original set of $n$ individual molecules in a dataset and hold out $m$ molecules (as
opposed to holding out a set of observations from the pairwise dataset). We then evaluate the SALI
values $(n-m)(n-m-1)/2$ pairs of molecules and develop a random forest model. In contrast to the
models previously described, we rebuilt these models using the entire set of $n$ molecules. The
model is then used to predict the SALI values between the $m$ holdout molecules and the $(n-m)$
remaining molecules.

We first performed this analysis with the Kalla dataset. Since predictions of new molecules involve
$n$ predictions per molecule, we removed three molecules from the dataset and rebuilt the random
forest model. The hold out molecules, were specifically selected since they displayed significant
activity cliffs with various members of the training set. We then evaluated the pairwise descriptor
values for these three with the training set molecules and obtained predictions of the log(SALI)
values. Figure \ref{fig:tset-pset-kalla}A summarizes the performance of the model on the training set
as well as for the prediction set. From a numerical point of view, the performance of the model on
he prediction set degrades somewhat (RMSE of 0.51 versus 0.41 for the training set). As noted above,
the variance of the prediction is higher at the lower scale of log(SALI) values. However, at the
higher end of the scale there a number of pairs that are well predicted. Yet, it is clear that a
number of relatively significant activity cliffs have been underestimated.

Given that large SALI values can arise due to high structural similarity, even when the activity
difference is small, simply examining log(SALI) values is not completely informative. Specifically,
we are interested in the predictions for the actual activity cliffs, where both the difference in
activity and the structural similarity is very high. Figure \ref{fig:tset-pset-kalla}B visualizes this
information. Each point corresponds to a pair of molecules, which are shaded by the absolute
prediction residual for that pair. Thus, the ``true'' cliffs are represented by points lying towards
the bottom right corner of the plot. In this case we see that there are a number of pairs of
molecules with $1 - T_C \leq 0.1$ and 100-fold or better difference in activity. While many of these
pairs are associated with a high prediction residual, there are a number with low to medium
residuals.

Figure \ref{fig:kalla-comparison} displays the structures of the three hold out molecules (top row)
and a selection of structures, which which these hold out molecules were predicted to exhibit an
activity cliff. For each molecule, the $K_i$ value is listed along with the absolute residual value
(log(SALI) units). For molecules \textbf{349288} and \textbf{349699}, the predicted cliffs are
relatively accurate, as evidenced by the low residuals. For \textbf{349138} on the other hand, most
of the residuals were were relatively high. While we have highlighted a number of cliffs,
Figure \ref{fig:tset-pset-kalla}C makes it clear that the predictions for this molecule were, on
average, worse than for the other two molecules.

Figure \ref{fig:tset-pset-dai} summarizes the results for the same set of analyzes described above,
applied to the Dai dataset. As before we removed three molecules (top row of
Figure \ref{fig:dai-comparison}) and rebuilt the random forest on the paired cases with the remaining
41 molecules. We then predicted the pairwise SALI values for the three hold-out molecules with the
remaining 41 molecules (i.e., the prediction set). A plot of the predicted versus observed log(SALI)
values for the training and prediction sets are displayed in Figure \ref{fig:tset-pset-dai}A. As
before the model performs poorly on the lesser cliffs, but improves for the more significant cliffs
(RMSE for the training set and prediction sets were 0.37 and 0.43 log units respectively). However,
the 2 most significant cliffs are under-estimated. However, in comparison to the Kalla dataset, the
distribution of absolute residuals aggregated by the hold-out molecule
(Figure \ref{fig:tset-pset-dai}C) are similar, with a median absolute residual of less than
0.3 log units.  For this dataset, the number of ``real'' cliffs is relatively low as shown in
Figure \ref{fig:tset-pset-dai}B. Interestingly, the compound pairs that display moderate to
significant cliffs are relatively well predicted - in fact, the maximum residuals are observed for
the least significant cliffs.

We then considered the three hold-out molecules and some of the members of the training set with
which they are predicted to show activity cliffs. For this dataset, there are relatively few
``true'' activity cliffs. For example, while molecules \textbf{371259} and \textbf{371307} have a
1000-fold difference in activity, their $T_c = 0.68$, and would not represent a truly significant
cliff. However, the pair is well predicted with a residual of 0.04 log units. if we consider the
subset of predictions where the $T_c \geq 0.8$ we note that the median absolute residual is 0.24 log
units. However, the activity differences are not always significant ranging from 1.1-fold to
13-fold. Two such examples are shown in Figure \ref{fig:dai-comparison}, where the only difference is
in the position of the methyl substituent. Similar behavior is observed with the other hold-out
molecules.

@@

Given that many of the models show relatively poor statistics on the pairwise data, it is important
to note that much of this is due to the large variance in the predictions for the less significant
cliffs. One possible reason for the increase in variance for these portions of the dataset is the
fact that the distribution of the log(SALI) values tends to be skewed to the right
(Figure \ref{fig:logsalidist}). In general we see that the smaller SALI values constitute a relatively
small portion of the dataset.

@@

\ctable[cap={}, caption={Performance metrics for the random forest
  models built on the single molecule activity data, for each ChEMBL dataset. In each case the model
was built using the logarithm of the reported activities.},
label={tab:rfmodel-orig}]
{lrrr}
{

}
{\FL
Dataset & Y-range & RMSE & $R^2$
\ML
Costanzo & 4.67 & 0.80 & 0.18 \NN
Kall          & 4.44 & 1.00 & 0.32 \NN
Dai          &  3.60 & 0.76 & 0.34
\LL
}

@@

\ctable[cap={}, caption={Three of the hold out molecules for the Kalla dataset
  and training set members with which the hold outs exhibit predicted
  activity cliffs. Bold numbers are ChEMBL MOLREGNO values and numbers
  in parentheses are the absolute prediction
  residual in log(SALI) units.}, figure, botcap,
label={fig:kalla-comparison}]    
{c}
{}
{\includegraphics[width=\linewidth]{kalla-comparison}}
